---
title: "Chapter 7.8"
author: "Ruben J, Hermann"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Hmsc)
library(ape)
library(MASS)
library(tidyverse)
library(bayesplot)
library(corrplot)
```


## 7.8 Simulated Case Studies with HMSC

### 7.8.1 Generating Simulated Data

```{r}
n = 200
ns = 5
X = cbind(rep(1, n), rnorm(n), rnorm(n))
beta1 = rep(0, ns)
beta2 = c(2,2,-2,-2,0)
beta3 = c(1,-1,1,-1,0)
beta = cbind(beta1, beta2, beta3)
L = X %*% t(beta)
Y = 1*((L + matrix(rnorm(n*ns), ncol = ns)) > 0)
```

In this script we have assumed that species 1 and 2 respond positively to the covariate $x_1$ whereas species 3 and 4 respond negatively. We have further assumed that species 1 and 3 respond positively to the covariate $x_2$ whereas species 2 and 4 respond negatively. Species 5 does not respond to either of the covariates.  
Note that when generating the data, we have assumed no residual species associations, meaning that the species occur independ￾ently of each other after accounting for their responses to the covariates $x_1$ and $x_2$.

```{r}
colMeans(Y)
```

### 7.8.2 Defining and Fitting Three Alternative HMSC Models

```{r}
XData = data.frame(x1 = X[,2], x2 = X[,3])
studyDesign = data.frame(sample = as.factor(1:n))
rL = HmscRandomLevel(units = studyDesign$sample)
models = list()
for (i in 1:3){
  XFormula = switch(i, ~1, ~x1, ~x1+x2)
  m = Hmsc(Y = Y, XData = XData, XFormula = XFormula,
  studyDesign = studyDesign, ranLevels = list
  (sample = rL),distr = "probit")
  models[[i]] = m
}
```

```{r}
nChains = 2
thin = 5
samples = 1000
transient = 500*thin
verbose = 500*thin
```
```{r}
for (i in 1:3){
  models[[i]] = sampleMcmc(models[[i]], thin = thin,
  samples = samples,transient = transient,
  nChains = nChains,verbose = verbose)
}
```

```{r}
mpost = convertToCodaObject(models[[3]])
ess.beta = effectiveSize(mpost$Beta)
psrf.beta = gelman.diag(mpost$Beta, multivariate = FALSE)$psrf
ess.omega = effectiveSize(mpost$Omega[[1]])
psrf.omega = gelman.diag(mpost$Omega[[1]], multivariate =FALSE)$psrf
```

Creating the histogram plots
```{r}
hist(ess.beta,ylab="Frequency",xlab=~paste("Effective sample size (parameter ",beta,")"),main=NULL)
hist(psrf.beta,ylab="Frequency",xlab=~paste("Potential scale reduction factor (parameter ",beta,")"),main=NULL)

hist(ess.omega,ylab="Frequency",xlab=~paste("Effective sample size (parameter ",Omega,")"),main=NULL)
hist(psrf.omega,ylab="Frequency",xlab=~paste("Potential scale reduction factor (parameter ",Oemga,")"),main=NULL)
```

### 7.8.3 Parameter Estimates in the HMSC Models

```{r}
for (i in 1:3){
  OmegaCor = computeAssociations(models[[i]])
  supportLevel = 0.95
  toPlot = ((OmegaCor[[1]]$support > supportLevel)
            + (OmegaCor[[1]]$support < (1-supportLevel)) > 0)
  + OmegaCor[[1]]$mean
  corrplot(toPlot, method = "color", col = c("grey","white","black"))
}
```

```{r}
partition = createPartition(m, nfolds = 2, column ="sample")
partition.sp = c(1,2,3,4,5)
result = matrix(NA, nrow = 3, ncol = 3)
for (i in 1:3){
    m = models[[i]]
    #Explanatory power
    preds = computePredictedValues(m)
    MF = evaluateModelFit(hM = m, predY = preds)
    result[1,i] = mean(MF$TjurR2)
    #Predictive power based on cross-validation
    preds = computePredictedValues(m, partition = partition)
    MF = evaluateModelFit(hM = m, predY = preds)
    result[2,i] = mean(MF$TjurR2)
    #Predictive power based on conditional cross-validation
    preds = computePredictedValues(m, partition = partition,
    partition.sp = partition.sp,
    mcmcStep = 100)
    MF = evaluateModelFit(hM = m, predY = preds)
    result[3,i] = mean(MF$TjurR2)
}
```

